Spatial Analysis

### 2021-06-15 Scripts/spatial_analisis.Rmd
better explained index

# Load data

## Maps data

@@

## Spatial component

@@

state_bf_plot <- ggplot() +
  geom_sf(data = eez_sf, aes(), fill = NA) +
  geom_sf(data = state_bf, aes(color = state), fill = NA)
#

@@

Here we expand a grid within the buffer so we can estimate the proportion of each state

@@

# Looks good
state_sf %>%
  st_transform(4326) %>% # to match shape
  st_set_crs(4326) %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 10000) %>%
  ggplot() +
  geom_sf(aes(color = state))+
  # geom_sf(data = eez_sf, aes(),fill =NA) +
  # ggplot()+
  geom_tile(data = ne_grid,
            aes(
              x = lon,
              y = lat
            ),
            alpha = 0.2
  )

@@

gridExtra::grid.arrange(
  # Overall (overlapping) position
  ggplot(grid_sf) +
    geom_sf(aes(color = state), alpha = 0.3) +
    geom_sf(data = eez_sf, aes(),fill =NA) + 
    theme(legend.position = ""),
  # Showing each state separatley
  ggplot() +
    geom_sf(data = grid_sf, aes(color = state),size = 0.1, alpha = 0.3) +
    facet_wrap(~state) +
    theme(legend.position = "top"),
  nrow = 1)

@@

## Interpolation rutine

Here we [interpolate](https://swilke-geoscience.net/post/spatial_interpolation/) the survey data within the previously created grid.

@@

### Function for interpolation

This is the main function used to interpolate the data per year. It follows a Triangular Irregular Surface method using the `interp::interp()` function.

@@

### Run function

@@

### Ocean adapt data

@@

@@

### Control Pannel

Load required data. Note that some of it has been previously created

@@

### Run routine

@@

```{r load_data}

@@

## Points

@@

ggplot(data_grid) +
  geom_tile(
    aes(
      x = lon,
      y = lat,
      fill = mean,
      col = mean
    )
  ) +
  facet_wrap(~ ID  +  region,
             ncol = 4) +
  scale_fill_viridis("Mean Proportion", alpha = 0.8) +
  scale_color_viridis("Mean Proportion", alpha = 0.8)

## Map

@@

state_fit <- historic_tif %>% 
  group_by(state,year,region) %>% 
  summarise(state_value = sum(value,na.rm= T), .groups = "drop") %>% 
  left_join(total_fited,
            by = c("year","region")) %>%
  mutate(percentage = state_value/total_value*100) %>% 
  left_join(periods) %>% 
  group_by(ID = state,period,region) %>% 
  summarise(mean_per = round(mean(percentage)),.groups = "drop")

@@

This graph shows the proportion of each State over time

@@

ggplotly(p,
         dynamicTicks = TRUE) %>% 
  layout(hovermode = "x") %>% 
  # add_trace() %>% 
  rangeslider()

```
## Area plot (running)

This graph shows the proportion of each State smoothed over a 10 years running mean. It helps seeing the trends better (I Think...)

>>>>

# Methods

## Spatial data preparation

### Load data

#### Maps data

@@

@@

state_bf_plot <- ggplot() +
  geom_sf(data = eez_sf, aes(), fill = NA) +
  geom_sf(data = state_bf, aes(color = state), fill = NA) +
  scale_color_manual(values = state_pallet)
#

@@

Here we expand a grid within the buffer so we can estimate the proportion of each state.

*Note:* Dark grey shaded area is the grid

@@

# Looks good
state_sf %>%
  st_transform(4326) %>% # to match shape
  st_set_crs(4326) %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 10000) %>%
  ggplot() +
  geom_sf(aes(color = state))+
  # geom_sf(data = eez_sf, aes(),fill =NA) +
  # ggplot()+
  geom_tile(data = ne_grid,
            aes(
              x = lon,
              y = lat
            ),
            alpha = 0.2
  ) +
  scale_color_manual(values = state_pallet)

@@

gridExtra::grid.arrange(
  # Overall (overlapping) position
  ggplot(grid_sf) +
    geom_sf(aes(color = state), alpha = 0.3) +
    geom_sf(data = eez_sf, aes(),fill =NA) + 
    scale_color_manual(values = state_pallet) +
    theme(legend.position = ""),
  # Showing each state separatley
  ggplot() +
    geom_sf(data = grid_sf, aes(color = state),size = 0.1, alpha = 0.5) +
    facet_wrap(~state) +
    theme(legend.position = "top") +
    scale_color_manual(values = state_pallet),
  nrow = 1)

@@

### Interpolation rutine

In this step we [interpolate](https://swilke-geoscience.net/post/spatial_interpolation/) the survey data within the previously created grid following a Triangular Irregular Surface method.

@@

#### Functions needed

We need to create a couple of functions to run the whole process

##### Interpolation main fx (tis).

This is the main function used to interpolate the data per year. It follows a Triangular Irregular Surface method using the `interp::interp()` function. If you want to see the function clic on `code`

@@

##### Run function

This is a sub-function that runs the `tis()` function by taxa and region. It saves the output as a .csv file for each species.

@@

#### Survey data

The interpolation was done with NOAA Northeast Fisheries Science Center Spring and Fall Bottom Trawl Surveys [data](https://www.fisheries.noaa.gov/region/new-england-mid-atlantic#science) provided by [Ocean adapt](https://oceanadapt.rutgers.edu/). Data was accessed trough the [Github](https://github.com/pinskylab/OceanAdapt).

*In primary publications using data from the database, please cite Pinsky et al. 2013. Marine taxa track local climate velocities. Science 341: 1239-1242 doi: 10.1126/science.1239352, as well as the original data sources.*

@@

##### Splitting up data

- No part on spatial analysis. Can be ignored.

This is just a sub-step to split up the data into single species files. This makes the app faster as it only needs to load species specific data, rather than all the data at de begining.

@@

#### Control Pannel

This is where we load the required data and prepare to run the interpolation function. Note that some of the data here has been previously created

@@

#### Run routine

Here we just run the routine for each of the species present in the Northeast US Fall and Spring surveys between 1973 and 2019.

- Note there are 43 identified taxa
- Some taxa do not have presence data in some years

@@

This section shows preliminary results for Black sea bass (*Centropristis striata*).

```{r load_data, eval = T, echo = T}

@@

## Average proportion

This map shows the average proportion of the extrapolated value from the whole study period within each State's water.

*Note:* Grey tiles represent `NAs`

@@

ggplot(data_grid) +
  geom_tile(
    aes(
      x = lon,
      y = lat,
      fill = mean,
      col = mean
    )
  ) +
  facet_wrap(~ ID  +  region,
             ncol = 4) +
  scale_fill_viridis("Mean Proportion", alpha = 0.8) +
  scale_color_viridis("Mean Proportion", alpha = 0.8)
```

## Proportion Change

Here we show the average proportion of the interpolation by State and time period. Time periods were arbitrary defined as;

- Early; 1972 to 1984
- Mid; 1985 to 1997
- Late; 1998 to 2011
- Now; 2012 to 2019

*Notes:* To panel represents Fall survey and bottom panel Spring. This computation considers the Overlapping of state waters.

@@

state_fit <- historic_tif %>% 
  group_by(state,year,region) %>% 
  summarise(state_value = sum(value,na.rm= T), .groups = "drop") %>% 
  left_join(total_fited,
            by = c("year","region")) %>%
  mutate(percentage = state_value/total_value*100) %>% 
  left_join(periods,
            by = "year") %>% 
  group_by(ID = state,period,region) %>% 
  summarise(mean_per = round(mean(percentage)),.groups = "drop")

@@

This graph shows the proportion of the interpolation value each State has over time.

*Note:* This plot is interactive. For ease comparison between States you can;

- *One* click on any State to *remove* it from the plot 
- *Two* clicks on any State to *isolate it* from the plot (other states can then be added by just clicking on them).
- The bottom panel allows you to reduce the time frame

@@

ggplotly(p,
         dynamicTicks = TRUE) %>% 
  layout(hovermode = "x") %>%
  rangeslider()

## Area plot (running mean)

This graph shows the proportion of each State smoothed over a *10 years running mean**. It allows to better see increasing and decreasing trends.

*Note:* This plot is also interactive and thus, follows the same options of the previous one.
